---
title: "Cyclistic Project"
author: "Andrew Pace"
date: "11/17/2021"
output: html_document
---

# Cyclistic Data Analysis Project

Cyclistic is a bike-sharing company in Chicago. For the purposes of this analysis, users fall into one of two categories: Cyclistic members and casual riders. Cyclistic members have purchased an annual membership, while casual riders have purchased either a single-ride pass or full-day pass. Cyclistic's goal is to convert casual riders into Cyclistic members. My task is to analyze how casual riders and Cyclistic members use the bikes differently. 

## Load libraries

```{r install packages}
library(tidyverse)
library(lubridate)
```

## Load CSV into dataframes

```{r load ride dataset}
ride_data <- read_csv("./data/edited_data/csv/combined_202011-202110.csv", 
    col_types = cols(ride_length = col_time(format = "%H:%M:%S"), 
        started_at = col_datetime(format = "%Y-%m-%d %H:%M:%S"), 
        ended_at = col_datetime(format = "%Y-%m-%d %H:%M:%S"), 
        start_station_id = col_character(), 
        end_station_id = col_character()))
```
```{r load weather dataset}
weather_data <- read.csv("./data/edited_data/csv/weather.csv")
```

## Data prep

I'll separate out the date and month from the columns started_at and ended_at. The new columns will be start_date, start_month, and end_date. This will make my plotting code cleaner and easier to read.

```{r separate date and time by creating new columns}
ride_data$start_date <- as.Date(ride_data$started_at)
ride_data$start_month <- month(ride_data$started_at,label = TRUE)
ride_data$end_date <- as.Date(ride_data$ended_at)
```

## Data exploration

### Number of rides of each user type

```{r number of members and casual}
casual_rider_count <- sum(ride_data$member_casual == 'casual')
member_rider_count <- sum(ride_data$member_casual == 'member')
total_rides <- casual_rider_count+member_rider_count
  
cat("Casual: ",casual_rider_count," rides (",casual_rider_count/total_rides*100,"%)\nMember: ",member_rider_count," rides (",member_rider_count/total_rides*100,"%)\nTotal: ", total_rides)
```

### Average ride length of each user type

```{r plot - avg members and casual}
casual_rider_avg <- time_length(mean(
  filter(ride_data, member_casual == "casual")$ride_length),"minute")
member_rider_avg <- time_length(mean(
  filter(ride_data, member_casual == "member")$ride_length),"minute")
cat("Casual: ",casual_rider_avg," minutes\nMember: ",member_rider_avg," minutes")
```

### Days of the week

```{r plot - day of the week}
plot_days_num <- ggplot(ride_data, aes(x=day_of_week, fill=member_casual))+
  geom_bar(position="dodge")+
  labs(
    title="Number of rides taken each day of the week",
    subtitle="Totaled over the course of a year",
    y="Number of Rides", 
    x="Day of the Week",
    caption="November 2020 - October 2021",
    fill="User Type")+
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))+
  scale_x_discrete(limits=c("1","2","3","4","5","6","7"),
                   labels = c(
                     "1"="Sunday",
                     "2"="Monday",
                     "3"="Tuesday",
                     "4"="Wednesday",
                     "5"="Thursday",
                     "6"="Friday",
                     "7"="Saturday"
                   ))+
  scale_fill_manual(values=c("#4E79A7", "#F28E2B"),
                       breaks=c("casual", "member"),
                       labels=c("Casual", "Member"))
plot_days_num
```

```{r plot - avg rides each day of week }
# Create new data frame to record averages for each day and user type
avg_ride_day <- data.frame(matrix(ncol = 3, nrow = 0))
# Name column headers
colnames(avg_ride_day) <- c('member_casual', 'day', 'avg')
days <- c(1:7)
# Loop through days
for (x in days){
  # Calculate average for that day for each member type
  mem_avg <- time_length(mean(
  filter(ride_data,
         member_casual == "member" &
           day_of_week == x)$ride_length),"minute")
  cas_avg <- time_length(mean(
  filter(ride_data,
         member_casual == "casual" &
           day_of_week == x)$ride_length),"minute")
  # Save date to data frame 
  avg_ride_day[nrow(avg_ride_day) + 1, ] <- c('member',x,mem_avg)
  avg_ride_day[nrow(avg_ride_day) + 1, ] <- c('casual',x,cas_avg)
}
# make the avg column numeric so the plot functions properly
avg_ride_day$avg <- as.numeric(avg_ride_day$avg)
plot_days_avg <- ggplot(avg_ride_day, aes(x=day, y=avg, fill=member_casual))+
  geom_bar(position="dodge", stat="identity")+
  labs(
    title="Average ride time each day of the week",
    subtitle="Averaged over the course of a year",
    y="Length of Ride (minutes)", 
    x="Day of the Week",
    caption="November 2020 - October 2021",
    fill="User Type")+
  scale_y_continuous(limits = c(0, 35), breaks = seq(0, 35, by = 5))+
  scale_x_discrete(limits=c("1","2","3","4","5","6","7"),
                   labels = c(
                     "1"="Sunday",
                     "2"="Monday",
                     "3"="Tuesday",
                     "4"="Wednesday",
                     "5"="Thursday",
                     "6"="Friday",
                     "7"="Saturday"
                   ))+
  scale_fill_manual(values=c("#4E79A7", "#F28E2B"),
                       breaks=c("casual", "member"),
                       labels=c("Casual", "Member"))
plot_days_avg
```

### Monthly rides

```{r plot - month}
plot_month_num <- ggplot(ride_data, aes(x=start_month, fill=member_casual))+
  geom_bar(position="dodge")+
  labs(
    title="Number of rides taken each month",
    y="Number of Rides", 
    x="Month",
    caption="November 2020 - October 2021",
    fill="User Type")+
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))+
  scale_fill_manual(values=c("#4E79A7", "#F28E2B"),
                       breaks=c("casual", "member"),
                       labels=c("Casual", "Member"))
plot_month_num
```

```{r plot - avg rides each month of year }
# Create new data frame to record averages for each month and user type
avg_ride_month <- data.frame(matrix(ncol = 3, nrow = 0))
# Name column headers
colnames(avg_ride_month) <- c('member_casual', 'month', 'avg')
months <- c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec')
# Loop through months
for (x in months){
  # Calculate average for that month for each member type
  mem_avg <- time_length(mean(
  filter(ride_data,
         member_casual == "member" &
           start_month == x)$ride_length),"minute")
  cas_avg <- time_length(mean(
  filter(ride_data,
         member_casual == "casual" &
           start_month == x)$ride_length),"minute")
  # Save date to data frame 
  avg_ride_month[nrow(avg_ride_month) + 1, ] <- c('member',x,mem_avg)
  avg_ride_month[nrow(avg_ride_month) + 1, ] <- c('casual',x,cas_avg)
}
# make the avg column numeric so the plot functions properly
avg_ride_month$avg <- as.numeric(avg_ride_month$avg)
# If I plotted now, the months would be in alphabetical order. 
# Reorder Months
avg_ride_month$month <- factor(avg_ride_month$month,levels=c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec'))
plot_month_avg <- ggplot(avg_ride_month, aes(x=month, y=avg, fill=member_casual))+
  geom_bar(position="dodge", stat="identity")+
  labs(
    title="Average ride time each month of the year",
    y="Length of Ride (minutes)", 
    x="Month",
    caption="November 2020 - October 2021",
    fill="User Type")+
  scale_y_continuous(limits = c(0, 35), breaks = seq(0, 35, by = 5))+
  scale_fill_manual(values=c("#4E79A7", "#F28E2B"),
                       breaks=c("casual", "member"),
                       labels=c("Casual", "Member"))
plot_month_avg
```

### Monthly temperature averages

To see if there is a correlation between rides and temperature I'll plot the monthly average temperatures recorded over the same time period.

```{r}
plot_temp_avg <- ggplot(weather_data, aes(month))+
  # geom_line(aes(y=min), color="blue")+
  geom_line(aes(y=avg), color="black")+
  # geom_line(aes(y=max), color="red")+
  labs(
    title="Average temperature each month of the year",
    y="Temperature (F)", 
    x="Month",
    caption="November 2020 - October 2021")+
  scale_y_continuous(limits = c(10, 80), breaks = seq(10, 80, by = 10))+
  scale_x_discrete(limits=
                     c("1","2","3","4","5","6","7","8","9","10","11","12"),
                   labels = 
                     c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec'))
plot_temp_avg
```

### Bike types

```{r plot - bike type}
plot_bike_num <- ggplot(ride_data, aes(x=rideable_type, fill=member_casual))+
  geom_bar(position="dodge")+
  labs(
    title="Number of rides taken on each type of bike available",
    y="Number of Rides", 
    x="Bike Types",
    caption="November 2020 - October 2021",
    fill="User Type")+
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))+
  scale_x_discrete(limits=c("classic_bike","docked_bike","electric_bike"),
                   labels = c(
                     "classic_bike"="Classic",
                     "docked_bike"="Docked",
                     "electric_bike"="Electric"
                   ))+
  scale_fill_manual(values=c("#4E79A7", "#F28E2B"),
                       breaks=c("casual", "member"),
                       labels=c("Casual", "Member"))
plot_bike_num
```


```{r plot - avg rides each type of bike }
# Create new data frame to record averages for each bike type and user type
avg_ride_bike <- data.frame(matrix(ncol = 3, nrow = 0))
# Name column headers
colnames(avg_ride_bike) <- c('member_casual', 'bike', 'avg')
bikes <- c('classic_bike','docked_bike','electric_bike')
# Loop through days
for (x in bikes){
  # Calculate average for that bike for each member type
  mem_avg <- time_length(mean(
  filter(ride_data,
         member_casual == "member" &
           rideable_type == x)$ride_length),"minute")
  cas_avg <- time_length(mean(
  filter(ride_data,
         member_casual == "casual" &
           rideable_type == x)$ride_length),"minute")
  # Save date to data frame 
  avg_ride_bike[nrow(avg_ride_bike) + 1, ] <- c('member',x,mem_avg)
  avg_ride_bike[nrow(avg_ride_bike) + 1, ] <- c('casual',x,cas_avg)
}
# make the avg column numeric so the plot functions properly
avg_ride_bike$avg <- as.numeric(avg_ride_bike$avg)

plot_bike_avg <- ggplot(avg_ride_bike, aes(x=bike, y=avg, fill=member_casual))+
  geom_bar(position="dodge", stat="identity")+
  labs(
    title="Average ride time on each type of bike available",
    y="Length of Ride (minutes)", 
    x="Bike Types",
    caption="November 2020 - October 2021",
    fill="User Type")+
  scale_y_continuous(limits = c(0, 55), breaks = seq(0, 55, by = 5))+
  scale_x_discrete(limits=c("classic_bike","docked_bike","electric_bike"),
                   labels = c(
                     "classic_bike"="Classic",
                     "docked_bike"="Docked",
                     "electric_bike"="Electric"
                   ))+
  scale_fill_manual(values=c("#4E79A7", "#F28E2B"),
                       breaks=c("casual", "member"),
                       labels=c("Casual", "Member"))
plot_bike_avg
```
